Efficiency

### 2018-01-15 DraftExec.tex
with all the indiscretion ratios

@@

\maketitle

Para

@@

\section{Introduction}

@@

\section{Executive Summary}

\subsection{The difference between nuclear and conventional for speed}
This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine. A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.  Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.  Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.

\subsection{The importance of low speed operation for diesel-electric submarines}
The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.  Consequently, excessive energy consumption results in greatly reduced dived endurances.  Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.  For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.  In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.

\subsection{What is cavitation}
Pumpjets have been widely adopted by navies operating nuclear submarines.  The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.  Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.

\subsection{How does a pump-jet work}
A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.  This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.  In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.  This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.

\subsection{The necessity of drag, and decelleration, induced by the duct}
The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.  By a fundamental requirements of physics, this actually requires that the water's incoming speed be \textbf{reduced} before it reaches the impeller. (The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller. Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.  Put simply, the water has to slow down to go fast again.

\subsection{What happens to a jet at low speed, in simple terms}
The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.  This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.  (Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system. Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.

\subsection{Why this isn't easily noticeable in normal circumstances}
It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).  This doesn't mean that they don't still work at low speeds and produce some thrust.  All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.

\subsection{What the submarine requirement tells us about the duct}
Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.  Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.  In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.  Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.

\subsection{Quantitative Conclusions}
In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine. The most plausible scenarios I find include the reduction of dived endrance and range between 20\% and 50\%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.

\subsection{The acoustic advantage of jets at higher speed}
My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation. Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.   As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.  This is known as a higher 'tactical silent speed'.

\subsection{Turbulence and flow separation at low speed}
However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.  Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.  These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.  In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.  However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.

\subsection{The acoustic question at low speed}
But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.  These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.  It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates. As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.  In a variety of circumstances, including most operations at patrol speed, it may not always be true.

\subsection{The inevitable trade-off}
However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.  These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.

\subsection{The contradiction between stated requirement and chosen technology}
Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.

\section{Speed and Drag - Why very slow is very very $(very)^2$ economical}

@@

\section{Some essential concepts}

\begin{figure}
\includegraphics[width=\textwidth]{BoundaryLayer.png}
\caption{The development of a boundary layer as shown in \parencite{mollard2011}}
\label{fig:BoundaryLayer.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{LaminarTurbulent.png}
\caption{A simple comparison of laminar and turbulent flows from \parencite{NPdotnet2017}}
\label{fig:LaminarTurbulent.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{EarlyCavitation.JPG}
\caption{Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs}
\label{fig:EarlyCavitation.JPG}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{BoilingCavitation.JPG}
\caption{Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface}
\label{fig:BoilingCavitation.JPG}
\end{figure}

\begin{figure}
\includegraphics[width=0.7\textwidth]{IncreasingCavitation.png}
\caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in \parencite{mollard2011}}
\label{fig:IncreasingCavitation.png}
\end{figure}

\begin{figure}
\includegraphics[width=0.7\textwidth]{CavitationTypes.png}
\caption{Different types of cavitation can occur on different parts of the propulsion system as shown in \parencite{carlton2007}}
\label{fig:CavitationTypes.png}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{SeparatedAerofoil.jpg}
\caption{Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)}
\label{fig:SeparatedAerofoil.jpg}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{FlowSeparation.png}
\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
\label{fig:FlowSeparation.png}
\end{figure}

Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.  Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.  This is the reason that golf balls have dimples, and tennis-balls have fur, to enable them to fly further.  The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.

\section{Basics of Ducted Propellers and Pumpjets}
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.  Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.

@@

In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.  Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it. \cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.  This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.

@@

\subsubsection{Waterjets}
The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.

\begin{figure}
\includegraphics[width=\textwidth]{PumpjetWaterjet.png}
\caption{A comparison of a pumpjet and waterjet as given in \cite{wislicenus1973}}
\label{fig:PumpjetWaterjet.png}
\end{figure}

\section{Constraints on the efficiency of pumpjets at low speed}

@@

\begin{figure}
\includegraphics[width=\textwidth]{BMTShape.png}
\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface. \parencite{giles2010}}
\label{fig:BMTShape.png}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{BMTEfficiency.png}
\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
\label{fig:BMTEfficiency.png}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{JetHighFlow.png}
\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red. After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
\label{fig:JetHighFlow.png}
\end{figure}

@@

Flow separation also tends to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.  As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.  These effects also have acoustic consequences, which will be discussed in Section \ref{Acoustic Considerations}.

\begin{figure}
\includegraphics[width=\textwidth]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{Comparison.png}
\caption{At the end of a tapering body the impact of the contracting hub will be much larger near the intake than near the tail, making the degree of diffusion ahead of the impeller, as well as contraction of the flow in the nozzle after it, much larger in impact than it appears by looking at the external shape. Credit \cite{gearhart1966selection}}
\label{fig:Comparison.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{Triomph.jpg}
\caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves. Credit \cite{reddit2016}}
\label{fig:Triomph.jpg}
\end{figure}

\subsection{Trade-offs and Limitations on Improvements}

@@

\begin{figure}
\includegraphics[width=\textwidth]{PumpjetEfficiency.png}
\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
\label{fig:PumpjetEfficiency.png}
\end{figure}

@@

Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.  In an idealised circumstance, the flow rate increases directly with vessel speed.  Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.  However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.  So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{figLPumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.  One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.

@@

\section{Model Structures and Assumptions}

@@

\section{Results and Discussion}

@@

>>>>

%  For caption width
\usepackage{graphicx,caption}
% For Centering figs
\usepackage[export]{adjustbox}
% For split figures
\usepackage{subcaption}
%  appendix
\usepackage[titletoc,page]{appendix}

@@

\begin{titlepage}
    \begin{center}
        \vspace*{1cm}

        \vspace{10cm}
        \textbf{\huge Efficiency}

        \vspace{0.5cm}

        \vspace{1.5cm}

        \textbf{Aidan Morrison}

        \vspace{1.5cm}
        2018

        \vfill

    \end{center}
\end{titlepage}

\newpage

@@

\newpage

\section{Introduction} \label{intro}

@@

\newpage  \section{Executive Summary}

The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.  Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.

Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."  No further explanation of this superiority has been provided in detail, but it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.

The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.  The conclusion of this report is that the following statement can be made with confidence:
\\
\\
\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.  This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
\\
\\
Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.  One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.  This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt. During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.  Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.  Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.

This report also explores in some detail the theoretical concepts which drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.  These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthmore, a particular hydrodynamic phenonomon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propuslor.  At very low speeds waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.  The act of slowing the water down is necessarily equivalent to the act of raising its pressure.  It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.  Maintaing this advantage to the same level necessitates a duct design which slows water down to the same extent.  The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.

In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.  The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems  \parencite{giles2010}.  Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modelled in all directions, and the combinations combined and tested.

The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the general public.

The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and known characteristics \ref{buckingham2008submarine}.  A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.  The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.  This advantage reached a proportional maxium at around 6kt of over 60\%.  However, in nominal terms, the greatest differences occurred at slightly lower speeds.  Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.

The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.  Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.  None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.  Given the significant emphasis that has been repeatedly placed the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about suitability of this technology to meet Australia's needs.

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitity of the difference to different assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

The model demonstrated clearly that all of the most pronounced advantages in dived range and endrance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.  The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.

The consequences of flow separation also has acoustic consequences, which this report explores.  The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.  This results in pressure fluctations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.  These effects are are likely to cause some accoustic signature which becomes lower in frequence as the submarine slows down further and the vortexes become larger, and shed more slowly.  This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequences as speed increases and bubbles expand.  Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.

It remains plausible that in the absence of cavitation the noise generated in the propulsor by such flow separation is not loud enough to be of significant tactical concern. However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.  Flow separtion demand that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust. Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.  This claim has strong grounds wherever a conventional propeller might experience cavitation, such as at higher speeds.  But at some very low speeds it is unlikely to be true.

Overall, it seems unlikely that the full range of consequences of pumpjet choice have been fully comprehended by the Australian government.  The scale of the probable impact on range and endurance is quite probably so substantial that it is difficult to see how such a performance penalty is consistent with the government's stated aim of acquiring a regionally superior submarine.

\newpage  \section{Speed and Drag - Why very slow is very very $(very)^2$ economical}

@@

\newpage  \section{Basics of Ducted Propellers and Pumpjets}
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.  Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.

@@

In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.  Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it. \cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.  This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in Section \ref{waterjets}.

@@

\subsubsection{Waterjets} \label{waterjets}

The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.

\begin{figure}[h]

\begin{subfigure}{0.5\textwidth}
\includegraphics[width=0.9\linewidth]{Pumpjet.png}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=0.9\linewidth]{Waterjet.png}
\caption{}
\end{subfigure}

\caption{A comparison of a pumpjet (a) and waterjet (b) as given in \cite{wislicenus1973}}
\label{fig:PumpjetWaterjet.png}
\end{figure}

\newpage  \section{Constraints on the efficiency of pumpjets at low speed}

@@

\begin{figure}[h]
\captionsetup{width=\textwidth}
\includegraphics[width=\textwidth]{BMTShape.png}
\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface. \parencite{giles2010}}
\label{fig:BMTShape.png}
\end{figure}

\begin{figure}

\begin{subfigure}[b]{0.9\textwidth}
\includegraphics[width=0.8\linewidth, center]{BMTEfficiency1.png}
\end{subfigure}
\begin{subfigure}[b]{0.9\textwidth}
\includegraphics[width=0.8\linewidth, center]{BMTEfficiency2.png}
\end{subfigure}

\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
\label{fig:BMTEfficiency.png}
\end{figure}

@@

@@

\begin{figure}[h]
\captionsetup{width=1\textwidth}
\includegraphics[width=1\textwidth]{JetHighFlow.png}
\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red. After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
\label{fig:JetHighFlow.png}
\end{figure}

@@

Flow separation also tends to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.  As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.  These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations}.

\begin{figure}[h]
\captionsetup{width=1\textwidth}
\includegraphics[width=1\textwidth]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{Comparison.png}
\caption{At the end of a tapering body the impact of the contracting hub will be much larger near the intake than near the tail, making the degree of diffusion ahead of the impeller, as well as contraction of the flow in the nozzle after it, much larger in impact than it appears by looking at the external shape. Credit \cite{gearhart1966selection}}
\label{fig:Comparison.png}
\end{figure}

\begin{figure}[h]
\captionsetup{width=0.8\textwidth}
\includegraphics[width=0.8\textwidth, center]{Triomph.jpg}
\caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves. Credit \cite{reddit2016}}
\label{fig:Triomph.jpg}
\end{figure}

\subsection{Trade-offs and Limitations on Improvements} \label{tradeoffs}

@@

\begin{figure}
\captionsetup{width=0.6\textwidth}
\includegraphics[width=0.7\textwidth, center]{PumpjetEfficiency.png}
\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
\label{fig:PumpjetEfficiency.png}
\end{figure}

@@

Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.  In an idealised circumstance, the flow rate increases directly with vessel speed.  Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.  However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.  So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency.png}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.  One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.

@@

\newpage  \section{Model Structures and Assumptions}

@@

\newpage  \section{Results and Discussion}

@@

Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.  They equivalently impact the total amount of energy available for consumption before recharging batteries while snorkeling. There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.  The nominal impact, however increases directly and linearly with the total energy stored.

The results in nominal terms are striking in their magnitude.  In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.  Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.  Even under the efficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type-214 operating without AIP.
